---
title: "Session4"
author: "Marco Tello"
date: "2025-03-26"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(tibble)
library(readr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Welcome to Session 4! So far, you’ve learned how to work with data in R,
create functions, and explore basic data manipulations. In this session,
you’ll step into the world of tidy data and the tidyverse. We’ll also
show you how to leverage ggplot2 for exploratory data analysis.

We will use a gene expression matrix where:

-   Column 1 = Gene name (e.g., ENSEMBL IDs).

-   Columns 2-8 = Expression counts for 7 samples.

-   Samples 1 to 3 belong to one condition.

-   Samples 4 to 7 belong to a second condition.

By the end of this session, you should be able to:

-   Understand what tidy data is and how it differs from non-tidy data.

-   Read data using tidyverse functions and compare tibbles vs. data
    frames.

-   Grasp the basics of ggplot2 syntax: data, aes(), and layers.

-   Reshape data for plotting using pivot_longer() and understand “wide”
    vs. “long” formats.

-   Filter lowly expressed genes with tidyverse.

-   Normalize counts using counts per million (CPM) from edgeR, then
    visualize with ggplot2.

Let’s jump in!

# 1. Tidy Data 

## 1.1 What is Tidy Data?

In R, tidy data is an approach to structuring your data so that:

-   Each variable is in its own column.

-   Each observation is a row.

-   Each type of observational unit is a separate table

Think of your dataset as a well-organized closet, with each shelf
dedicated to a single item type—no mismatched socks stuffed in random
places! This structure makes it easier to perform manipulations,
analyses, and visualizations. 1.2 Advantages of Tidy Data

-   Consistency: Because each column is a clear variable, you always
    know where to find what you need.

-   Compatibility: Many R packages (especially within the tidyverse)
    expect data in tidy format to work smoothly.

-   Readability: It’s more intuitive to understand the data at a glance.

### 1.3 Example: 

## Tidy vs. Non-Tidy Non-Tidy Example

## Gene Sample1_expression Sample2_expression Condition1 Condition2

ENSG... 100 150 "blue" "red" ENSG... 200 300 "blue" "red"

Each column combines what should be separate variables
(Sample1_expression lumps “Sample” and “Expression” in the column name).
Conditions might be in separate columns, which could be better
represented as a single “Condition” variable.

### Tidy Example

## Gene Sample Condition Expression

ENSG... 1 "blue" 100 ENSG... 2 "blue" 150 ENSG... 1 "red" 200 ENSG... 2
"red" 300

Now you have one column for Gene, one for Sample, one for Condition, and
one for the Expression value. This layout is more flexible for functions
like ggplot(), dplyr verbs, etc.

# 2. Reading Data with the Tidyverse

The tidyverse is a collection of packages designed with shared ideas
about how data should be structured and manipulated (e.g., dplyr, tidyr,
readr, ggplot2).

You can install the tidyverse with:

```{r}
# install.packages("tidyverse")
library(tidyverse)

```

## 2.1 Tibble vs. Data Frame

When you read data with read_csv() (from readr, a core tidyverse
package), you get a tibble instead of a base R data frame. A tibble is
basically a data frame with:

-   A more refined print method (so it doesn’t flood your console with
    too many columns/rows).

-   Better handling of column names and data types.

### Advantages of Tibbles:

-   Better defaults: They don’t convert strings to factors by default.

-   Partial row printing: Tibbles show you a preview without overloading
    the console.

-   Implicit type-stable: You’re less likely to run into surprises
    during data manipulation.

-   You can still work with tibbles using all the regular data frame
    functions, so it’s generally a smooth transition.

# 3. ggplot2: Grammar of Graphics

ggplot2 is like constructing a plot by layering components:

-   data = The dataset you want to visualize.

-   aes() = Aesthetics that map columns to visual features (e.g., x, y,
    color, fill).

-   geom\_\* = Geoms or geometry layers that define how the data is
    drawn (points, lines, bars, densities, etc.).

### Basic Syntax:

ggplot(data = <DATA>, aes(<MAPPINGS>)) + \
<GEOM_FUNCTION>()

<DATA> is typically a data frame or tibble.

<MAPPINGS> are aes(x = something, y = something_else, color = group) or
similar.

Each + adds a new “layer” to the plot.

## 3.1 Example: Creating a Box Plot

Let’s create a quick example tibble to demonstrate. Suppose we have a
small dataset of Gene, Sample, and Expression:

```{r}
# Create a toy tibble
example_data <- tibble(
  Gene = c("ENSG0001", "ENSG0002", "ENSG0003", "ENSG0004"),
  Sample1 = c(100, 150, 300, 500),
  Sample2 = c(80, 130, 350, 400)
)

# For ggplot, we often want the data in LONG format:
long_data <- example_data %>%
  pivot_longer(
    cols = starts_with("Sample"), 
    names_to = "Sample",
    values_to = "Expression"
  )

# Now let's plot a boxplot:
ggplot(long_data, aes(x = Sample, y = Expression)) +
  geom_boxplot() +
  labs(title = "Example Box Plot", x = "Sample", y = "Expression Value")

```

Here, we used:

pivot_longer() to go from wide to long.

aes(x = Sample, y = Expression) to map columns to axes.

geom_boxplot() to draw the boxplot layer.

# 4. Formatting a Tibble for ggplot: Pivoting

We often have expression data in wide format, something like:

Gene Sample1 Sample2 Sample3 Sample4 ... ENSG0001 100 120 130 80 ...
ENSG0002 400 200 300 500 ... ... ... ... ... ... ...

But ggplot2 loves long format, which might look like:

Gene Sample Expression ENSG0001 Sample1 100 ENSG0001 Sample2 120
ENSG0001 Sample3 130 ENSG0002 Sample1 400 ENSG0002 Sample2 200 ... ...
...

To get there, we use pivot_longer() from tidyr.

```{r, eval=FALSE}
long_data <- original_data %>%
  pivot_longer(
    cols = -Gene,           # or specify e.g. 2:8 if you know your gene col is 1
    names_to = "Sample",
    values_to = "Expression"
  )

```

## 4.1 Practice Exercises

Plot a Boxplot: From that toy data, create a simple boxplot of
Expression by Sample.

```{r}

# Toy data: 10 genes, 3 samples
example_data <- tibble(
  Gene = paste0("ENSG0000", 1:10),
  Sample1 = sample(50:500, 10),
  Sample2 = sample(50:500, 10),
  Sample3 = sample(50:500, 10)
)

```

# Putting it all together!

Let’s read a count matrix named count_df with columns:

ENSEMBL = Gene IDs (like ENSG000000...) in the first column.

Sample1 through Sample7 = Expression counts in columns 2 through 8.

```{r}

# Read a TSV file into a tibble
count_df <- read_tsv("~/GitHub/WorkshopR/IntroductoryLevel/data/exp_matrix.txt")

# Check the first rows
head(count_df)
```

## 5.1 Pivoting & Exploratory Plots

```{r}
count_df %>%
  pivot_longer(!ENSEMBL) %>%
  # Check if we want to remove the gene column or not
  dplyr::select(!ENSEMBL) %>%
  ggplot(aes(x = value, color = name)) +
    geom_density() +
    scale_x_continuous(trans = 'log10') +
    labs(title = "Gene expression density plot")

```

In this code:

-   pivot_longer(!ENSEMBL) means pivot all columns except the ENSEMBL
    column.

-   dplyr::select(!ENSEMBL) is removing the gene column from the final
    dataset for plotting. (Whether you need this line is
    context-dependent.)

-   We then call ggplot(), mapping x = value (the expression) and color
    = name (the sample).

## 5.2 Annotating Conditions

Next, we can add a “condition” column based on which sample belongs to
which group. For instance, samples 1 to 3 are “blue,” and samples 4 to 7
are “red.”

```{r}
temp <- count_df %>%
  pivot_longer(!ENSEMBLID, names_to = "name", values_to = "value") %>%
  mutate(sample_number = str_replace(string = name, 
                                     pattern = "Sample", 
                                     replacement = "")) %>%
  mutate(sample_number = as.numeric(sample_number)) %>%  # Convert to numeric
  mutate(condition = case_when(
    sample_number <= 3 ~ "blue",
    sample_number > 3  ~ "red"
  )) %>%
  dplyr::select(-sample_number)
 
temp %>%
  ggplot(aes(x = value, color = name)) +
  geom_density() +
  scale_x_continuous(trans = 'log10', labels = scales::comma) +
  scale_color_manual(values = temp$condition) +
  labs(title = "Gene expression density plot by condition")

```

We’ve introduced a condition column by using case_when(). Notice we also
used scale_color_manual() to assign custom colors.

## 5.3 Filtering Lowly Expressed Genes and using groups

Before diving further, you might want to remove genes with very low
expression across all samples. For example, keep only genes that have at
least 5 counts in half the samples.

```{r}
# Number of sample columns
num_samples <- ceiling(7 / 2)
threshold   <- 5

count_df_filtered <- count_df %>%
  # For each row (gene), count how many sample columns have counts >= threshold
  rowwise() %>%
  mutate(n_above_threshold = sum(c_across(starts_with("Sample")) >= threshold)) %>%
  # Keep rows where at least half the samples (rounded up) pass the threshold
  filter(n_above_threshold >= num_samples) %>%
  ungroup()

count_df_filtered

```

In the tidyverse, the concept of groups is central to how functions like
summarize(), mutate(), and filter() can be applied in a segmented way.
Here’s the big picture:

```         
group_by(): Creates “subsets” of your data based on one or more columns.

Subsequent Operations: Functions like summarize(), mutate(), or even filter() will then operate within each group independently.

Ungrouping: If you want to return to the dataset without these group boundaries, you can use ungroup().
```

```{r}


# Pretend this is pivoted data: each row is (Gene, Sample, Condition, Expression)
toy_long <- tibble(
  Gene       = rep(paste0("ENSG000", 1:5), each = 2),
  Sample     = rep(c("Sample1", "Sample2"), times = 5),
  Condition  = rep(c("blue", "red"), times = 5),
  Expression = c(100, 200, 150, 250, 120, 300, 85, 210, 400, 550)
)

# 1) Group by Condition
toy_long_grouped <- toy_long %>%
  group_by(Condition)

# 2) Summarize within each Condition
toy_summary <- toy_long_grouped %>%
  summarize(
    mean_expr = mean(Expression),
    sd_expr   = sd(Expression),
    count_genes = n()
  )

toy_summary
```

### Why Grouping Matters

```         
Clarity: You can explicitly define how data is segmented (e.g., by condition, timepoint, replicate).

Modularity: Summaries and statistics are produced for each subgroup.

Scalability: Works seamlessly with large datasets; no need to manually split data frames or write loops.
```

By combining group_by() with summarize(), mutate(), and even filter(),
you gain flexible, readable ways to handle complex biological
datasets—whether you’re computing means of expression by condition,
trimming out lowly expressed genes in specific groups, or normalizing
each group differently.

## 5.4 Normalizing Counts with CPM

Before further analysis, we often normalize gene counts to account for
different library sizes. A common approach is to convert counts to
counts per million (CPM), then transform to log space for better
visualization. We can do this with the edgeR package:

```{r}
# install.packages("edgeR") # Or use BiocManager::install("edgeR") if needed
library(edgeR)

# Suppose 'exp_mat' is your matrix or data frame of counts
# (excluding the gene names column).
# We'll do:
expr_log2cpm <- cpm(exp_mat, log = TRUE, prior.count = 1)

# Now let's pivot that data for plotting:
expr_log2cpm %>%
  as.data.frame() %>%
  rownames_to_column("ENSEMBLID") %>%   # If row names are gene IDs
  pivot_longer(!ENSEMBLID, names_to = "Sample", values_to = "Expression") %>%
  ggplot(aes(x = Expression, color = Sample))+
  geom_density() +
  labs(title = "Log2-CPM Normalized Gene Expression Density")

```

Key Points:

```         
cpm() computes CPM from raw counts.

log = TRUE applies a log2 transform to the CPM values.

prior.count = 1 avoids taking log of zero.

We then reshape the data with pivot_longer() to get a tidy format for ggplot.
```

## 5.5 Violin Plots

Finally, you can encourage your lab members to create violin plots on
their own—both for the raw counts and the normalized data. A violin plot
shows the distribution shape of the data, similar to a boxplot with a
rotated density on each side.

Hints: look into the aes() fill parameter and the function
scale_fill_manual() if you want to change the colors.

Conclusion

In this session, you learned:

```         
What tidy data is and why it’s beneficial.

How to use the tidyverse to read data into tibbles and the advantages tibbles offer.

The “grammar of graphics” in ggplot2, layering transformations via pivot_longer(), mapping aesthetics, and adding geometry layers.

How to filter out lowly expressed genes using tidyverse syntax.

How to use edgeR to perform CPM normalization on raw counts before visualizing.
```

With these skills, you’re ready to handle broader and more complex
explorations of gene expression data. Tidy formats will make your life
easier, and ggplot2 will give you the power to communicate results
visually. Happy coding!
